---
title: "rnnpsycholing NP/Z transitivity"
output:
  pdf_document: default
  html_notebook: default
---

We are looking for a main effect of ambiguity at the main verb.

Why is this interesting?
It indicates the NP/Z ambiguity is modulated by information about verb transitivity.

```{r}
rm(list = ls())
library(tidyverse)
library(lme4)
library(lmerTest)
library(plotrix)

REGION_ORDER = c("Start", "Embedded Verb", "Blocker", "Comma", "NP/Z", "Extension", "Verb", "Rest", "End")
REGION_EXEMPLARS = c("When the dog", "struggled/scratched", "his owner", ",", "the vet", "with his new assistant", "took off", "the muzzle", ". <eos>")
NUM_REGIONS = length(REGION_ORDER)

REGION_ORDER_NOBLOCKER = c("Start", "Embedded Verb", "Comma", "NP/Z", "Extension", "Verb", "Rest", "End")
REGION_EXEMPLARS_NOBLOCKER = c("When the dog", "struggled/scratched", ",", "the vet", "with his new assistant", "took off", "the muzzle", ". <eos>")
NUM_REGIONS_NOBLOCKER = length(REGION_ORDER_NOBLOCKER)

REGION_ORDER_NOCOMMA = c("Start", "Embedded Verb", "Blocker", "NP/Z", "Extension", "Verb", "Rest", "End")
REGION_EXEMPLARS_NOCOMMA = c("When the dog", "struggled/scratched", "his owner", "the vet", "with his new assistant", "took off", "the muzzle", ". <eos>")
NUM_REGIONS_NOCOMMA = length(REGION_ORDER_NOCOMMA)

rename_LSTM = function(d) {
  d %>% mutate(LSTM=if_else(LSTM == "gulordava", "GRNN", 
                    if_else(LSTM == "google", "JRNN",
                    if_else(LSTM == "rnng", "RNNG", LSTM))))
}


add_numeric_predictors <- function(d) {
  d %>%
    mutate(transitivity.numeric=if_else(transitivity=="transitive",1,-1),
           length.numeric=if_else(length == "short", -1, 1),
           comma.numeric=if_else(comma == "comma", 1, -1),
           blocker.numeric=if_else(blocker == "blocker", 1, -1))
           
}

d = read_csv("tests/combined_results_with_rnng_and_tinylstm.csv") %>%
  #select(-X1) %>%
  separate(condition, sep="_", into=c("transitivity", "length", "comma", "blocker")) %>%
  mutate(region=if_else(region == "Intransitive Verb", "Embedded Verb", region)) %>%
  mutate(region=if_else(region == "Transitive Verb", "Embedded Verb", region)) %>%
  mutate(region=factor(region, levels=REGION_ORDER)) %>%
  rename(LSTM=model) %>%
  mutate(unk=unk=="True",
         final=final=="True")


d_agg = d %>% 
  rename_LSTM() %>% 
  group_by(sent_index, region, transitivity, length, comma, blocker, LSTM) %>%  # get sum surprisal per region
    summarise(surprisal=mean(surprisal),
              unk=any(unk)) %>%
    ungroup() %>% 
  mutate(blocker=factor(blocker, levels=c("noblocker", "blocker")),
         transitivity=factor(transitivity, levels=c("transitive", "intransitive"))) %>%
  filter(!unk) 


dc = d_agg %>% filter(region == "Verb") %>% add_numeric_predictors() # critical region

```

## Overall visualization

```{r}

d_by_region = d_agg %>% 
  group_by(LSTM, region, sent_index) %>%
    mutate(item_mean=mean(surprisal)) %>%
    ungroup() %>%
  group_by(LSTM, region, transitivity, length, comma, blocker) %>%
    summarise(m=mean(surprisal),
              s=std.error(surprisal-item_mean),
              upper=m + 1.96*s,
              lower=m - 1.96*s) %>%
    ungroup() 

d_by_region %>%
  filter(length == "long", blocker == "noblocker") %>%
  mutate(region=factor(region, levels=REGION_ORDER_NOBLOCKER),
         region=as.numeric(region)) %>% 
  mutate(comma=if_else(comma == "nocomma", "no comma", "comma"),
         comma=factor(comma, levels=c("no comma", "comma"))) %>%
  ggplot(aes(x=region, y=m, ymax=upper, ymin=lower, color=transitivity, linetype=comma)) +
    geom_line() +
    geom_errorbar(linetype="solid", width=.1) +
    scale_x_continuous(breaks=seq(1, NUM_REGIONS_NOBLOCKER), labels=REGION_EXEMPLARS_NOBLOCKER) +
    theme_bw() +
    theme(axis.text.x = element_text(angle=15, hjust=1)) +
    xlab("") +
    ylab("Mean surprisal in region") +
    labs(color="", linetype="") +
    facet_grid(LSTM~., scale="free_y") +
    theme(legend.position="top")

ggsave("npz_transitivity.pdf", width=6, height=5)
  
```

## Plot the size of the garden path effect

```{r}
d_effect = d_agg %>%
  filter(region == "Verb") %>%
  filter(length == "long", blocker == "noblocker") %>%
  spread(comma, surprisal) %>%
  mutate(effect=nocomma-comma) %>%
  group_by(sent_index, LSTM) %>%
    mutate(item_mean=mean(effect)) %>%
    ungroup() %>%
  group_by(transitivity, LSTM) %>%
    summarise(m=mean(effect),
              s=std.error(effect - item_mean),
              upper=m+1.96*s,
              lower=m-1.96*s) %>%
    ungroup()

d_effect %>%
  ggplot(aes(x=transitivity, y=m, ymin=lower, ymax=upper, fill=transitivity)) +
    geom_bar(stat="identity", position="dodge") +
    geom_errorbar(width=.5) +
    facet_wrap(~LSTM) +
    ylab("Garden path effect (bits)") +
    xlab("Embedded verb transitivity") +
    theme_bw() +
    theme(legend.position="none")

ggsave("npz_effect.pdf", width=4, height=4)

```



## Is there an NP/Z garden path?

```{r}

mj.l = lmer(surprisal ~ comma.numeric + (1|sent_index), 
          data=filter(dc, LSTM == "JRNN", length == "long", blocker == "noblocker"))
mg.l = lmer(surprisal ~ comma.numeric + (1|sent_index), 
          data=filter(dc,  LSTM == "GRNN", length == "long", blocker == "noblocker"))
mr.l = lmer(surprisal ~ comma.numeric + (1|sent_index), 
          data=filter(dc, LSTM == "RNNG", length == "long", blocker == "noblocker"))
mt.l = lmer(surprisal ~ comma.numeric + (1|sent_index), 
          data=filter(dc, LSTM == "tinylstm", length == "long", blocker == "noblocker"))


summary(mj.l)
summary(mg.l)
summary(mr.l)
summary(mt.l)


```


## Is there an interaction of length and ambiguity?

Positive interaction = digging in effect, negative interaction = anti-digging in

```{r}

mji.l = lmer(surprisal ~ comma.numeric * transitivity.numeric + (comma.numeric + transitivity.numeric|sent_index), 
          data=filter(dc, LSTM == "JRNN", length == "long", blocker == "noblocker"))
mgi.l = lmer(surprisal ~ comma.numeric * transitivity.numeric + (comma.numeric + transitivity.numeric|sent_index), 
          data=filter(dc,  LSTM == "GRNN", length == "long", blocker == "noblocker"))
mri.l = lmer(surprisal ~ comma.numeric * transitivity.numeric + (comma.numeric + transitivity.numeric|sent_index), 
          data=filter(dc, LSTM == "RNNG", length == "long", blocker == "noblocker"))
mti.l = lmer(surprisal ~ comma.numeric * transitivity.numeric + (comma.numeric + transitivity.numeric|sent_index), 
          data=filter(dc, LSTM == "tinylstm", length == "long", blocker == "noblocker"))


summary(mji.l)
summary(mgi.l)
summary(mri.l)
summary(mti.l)



```

Nothing in JRNN, but anti-digging-in in GRNN.


